\textbf{Development of outdoor air pollution models for estimating exposure in the Barcelona Life Study Cohort (BiSC)}

### 2023-12-01 manuscript_BiSC_outdoor_AP.tex
Updates from Overleaf

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{authblk} % this is for multiple autho  rs 
\usepackage[colorlinks=true, allcolors=black]{hyperref}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{multirow}

@@

Air pollution stands as the primary environmental risk factor for human health and the fourth most deadly health risk worldwide, playing a significant role in the global burden of disease \cite{cohen2017, he2020state}. Exposure to ambient air pollution has been linked with a extensive array of adverse health outcomes \cite{boogaard2022, guxens2022hei, haddad2023}. In this context urbanization have been playing a major role increasing air pollution exposure as cities grow and more population moves to urban areas \cite{nieuwenhuijsen2016}. This trends present challenges for epidemiological studies \cite{tonne2017}, where accurate estimates capturing both the spatial and temporal variability are essential to evaluate the health impacts of air pollution in urban areas \cite{boogaard2022}.

Over the past few decades, several modelling techniques have been adopted in environmental epidemiology to determine exposure concentrations of air pollution \cite{hoek2017, di2019no2, di2019pm25, stafoggia2019, stafoggia2020}. Among these, Land use regression (LUR) and dispersion models (DM) are the most common approaches for estimating air pollution exposure concentrations in epidemiological studies \cite{gulliver2015, dehoogh2014}. LUR models employ regression techniques to predict air pollution levels at non-monitored locations, drawing on observed air pollution data and spatial explanatory variables that work as proxies for traffic-related air pollution \cite{briggs1997, hoek2008}. In contrast, DMs rely on mathematical simulations to estimate air pollution levels. These predictions are based on data from emissions sources, chemical transformation models, pollutant dispersion models, topography, roughness, and meteorological data \cite{hoek2017}. Although these methods have been extensively used they come with inherent limitations related to lack of temporal and spatial variation. To overcome the shortcomings, hybrid models (HM) have emerged, blending both spatial and temporal advantages, leading to refined and enhanced estimations \cite{hoek2017, korek2017, tularam2021, oh2021}.

While significant advancements have been made in spatiotemporal modelling of air pollution over large-scale domains, such models (multi-country and national domain models) often compromise their accuracy at smaller scales \cite{dehoogh2016, chen2019, chen2020, shen2022}. This gap underscores the need for a comprehensive assessment of outdoor air pollution exposure in smaller domains to understand the health impacts of air pollution exposure, especially during critical stages of development, due to the potential lasting effects of air pollution over the lifespan \cite{selevan2000, wick2010, ghosh2021}.

Urban areas add a layer of complexity to air pollution modelling due to complex spatiotemporal patterns that difficult accurate estimation of ground-level air pollutant concentrations \cite{sokhi2022}. In response to these challenges, machine learning (ML) algorithms have been increasingly used, particularly decision trees-based models \cite{liu2022}. This family of models is often used due to its ability to model potential, nonlinear interaction between predictor variables \cite{liu2022treebased}. Among the different ML algorithms, random forest (RF) has emerged as one of the most robust techniques for predicting outdoor air pollution concentrations \cite{chen2019, chen2020, stafoggia2019, stafoggia2020, schneider2020, mila2023}.

Furthermore, the COVID-19 pandemic has introduced unprecedented challenges for air pollution modelling and exposure assessment. These challenges encompass changes in air pollution patterns, constraints on data collection due to health and safety restrictions (lockdowns), and the complexity of data integration due to changes in traffic, mobility, and land use dynamics in cities \cite{gonzalez2022, querol2021}.

@@

Barcelona (\textbf{\cref{fig1}}), is one of the most populated and dense cities in Spain, with 1,656,725 inhabitants ($\approx16,255.4$ inhabitants/km\textsuperscript{2}). It is located on the Mediterranean coast of the Iberian peninsula, protected by the Collserola mountain range and delimited by two delta rivers (Besòs and LLobregat). The city has a Mediterranean climate, the wind patterns are dominated by a breeze that blows from the sea during the daytime and from the land during night-time. Barcelona is one of the cities in Europe with the highest vehicle traffic density (6,000 vehicles/km\textsuperscript{2}) \cite{casallas2018}.

@@

Air pollution measurements were conducted during two separate monitoring campaigns. The first, named the "BiSC-home" campaign, took place at the home addresses of BiSC participants and involved measuring only NO\textsubscript{2} levels for one week during the first, and third trimester of pregnancy between 2018-2021. The second campaign was carried out in accordance with the ESCAPE guidelines, named "BiSCAPE", and included four distinct measurement periods: three in the winter, summer, and autumn of 2021, and a final one in the winter of 2022. During these periods, levels of NO\textsubscript{2}, PM\textsubscript{2.5}, and BC were measured at 34 locations across Barcelona Metropolitan Area (Barcelona, Esplugues de Llobregat, and L'Hospitalet).These monitoring campaigns were used in the development of the LUR and HM.

@@

LUR models for NO\textsubscript{2}  were developed using only data from the locations that had measured home-outdoor (BiSC-home campaign) levels for both sampling campaigns (i.e., we did not use data from participants who had data available for only one campaign, either first or third trimester). Almost half of the sites were measured before the COVID-19 pandemic started and half of the sites were during the pandemic period and its resulting restrictions. For PM\textsubscript{2.5}, BC  and PM\textsubscript{2.5} constituents models were developed using BiSCAPE data from the 34 sites.

For the development of the LUR models, spatial variables that better represent the air pollutant concentration distribution across the study area were created. Following the ESCAPE guidelines \cite{eeftens2012, beelen2013},  predictor variables were selected from three broad categories: traffic, land use, and urban configuration. A wide variety of buffer sizes were created, i.e, 25m, 50m, 100m, 300m, 500m, and 1000m to capture the area-specific predictor variable at every single location. Predictor variables were calculated using specific coordinates for each site.

We applied a supervised stepwise process in which we sequentially added a predictor variable to a linear regression model using pollutant concentrations as the independent variable to obtain the best possible model \cite{eeftens2012, beelen2013}. This procedure was followed to build all the specific pollutant models. Previously to model development, we specified the direction expected of the effect that each predictor has in the model. For all predictor variables, univariate regression models with annual average air pollution concentration (NO\textsubscript{2}, PM\textsubscript{2.5}, BC, and PM\textsubscript{2.5} constituents (Fe, Cu, Zn)) were conducted, models with the highest adjusted explained variance (adjusted R\textsuperscript{2}) and the right direction of the effect were selected. We then repeated the process using the previous model and all the remaining predictor variables. Details of the LUR modelling and variables used can be found in the \textbf{Supplementary materials} .

@@

\subsubsection{LUR model evaluation}
Performance of the LUR models was assessed using \textit{k-fold cross-validation}. Specifically, we employed a 10-fold cross-validation technique. In this approach, the data was split into ten distinct sub-samples. For each iteration, the model was trained on k-1 folds and then validated using the leftover fold \cite{ziegel2003}. For every fold, we computed metrics such as \(R^2\), the root-mean-square-error (RMSE), mean bias (MB), and the correlation coefficient (\(r\)). After each iteration, both the error and adjusted \(R\textsuperscript{2}\) were recorded. This process was repeated \(k-times\), and the average of these measures was reported as the indicator of the model validity. 
\vspace{0.5cm} Details from the metrics used for model validation can be found in the \textbf{Supplementary material X}.

@@

\subsection{Dispersion model evaluation}
Dispersion models were evaluated by comparing the yielded results from the ADM-Urban model with the real values measured by the air quality stations (XVPCA), including Ciutadella, Eixample, Gracia, Palau Reial, Poblenou, Sants, among others. Details from the metrics used for dispersion model validation can be found in the \textbf{Supplementary material X}.\vspace{0.5cm}

@@

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%
%%% --- Results --- %%%
%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

@@

%%% --- Figure. Comparison/agreement between moodel predictions estimates --- %%%
% Put the figure text closer to the Figure 3
\captionsetup[figure]{skip=6pt}
% We add the figure with the study domain 
\begin{figure}[!htb]
\includegraphics[width=1.0\textwidth]{figures/NO2_HEX.png}
\end{figure}

% Put the figure text closer to the Figure 3
\captionsetup[figure]{skip=6pt}
% We add the figure with the study domain 
\begin{figure}[!htb]
\includegraphics[width=1.0\textwidth]{figures/PM25_HEX.png}
\end{figure}

% Put the figure text closer to the Figure 3
\captionsetup[figure]{skip=6pt}
% We add the figure with the study domain 
\begin{figure}[!htb]
\includegraphics[width=1.0\textwidth]{figures/BC_HEX.png}
\caption{Comparison/agreement between outdoor air pollution model estimates \textit{NO$_2$}, \textit{PM$_{2.5}$}, \textit{BC}.}
\end{figure}

>>>>

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{authblk} % this is for multiple autho  rs 
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{multirow}

@@

Air pollution stands as the primary environmental risk factor for human health and the fourth most deadly health risk worldwide, playing a significant role in the global burden of disease \cite{cohen2017, he2020state}. Exposure to ambient air pollution has been linked with a extensive array of adverse health outcomes \cite{boogaard2022, guxens2022hei, haddad2023}. In this context urbanization have been playing a major role increasing air pollution exposure as cities grow and more population moves to urban areas \cite{nieuwenhuijsen2016}. This trends present challenges for epidemiological studies \cite{tonne2017}, where accurate estimates capturing both the spatial and temporal variability are essential to evaluate the health impacts of air pollution in urban areas \cite{boogaard2022}. \\

Over the past few decades, several modelling techniques have been adopted in environmental epidemiology to determine exposure concentrations of air pollution \cite{hoek2017, di2019no2, di2019pm25, stafoggia2019, stafoggia2020}. Among these, Land use regression (LUR) and dispersion models (DM) are the most common approaches for estimating air pollution exposure concentrations in epidemiological studies \cite{gulliver2015, dehoogh2014}. LUR models employ regression techniques to predict air pollution levels at non-monitored locations, drawing on observed air pollution data and spatial explanatory variables that work as proxies for traffic-related air pollution \cite{briggs1997, hoek2008}. In contrast, DMs rely on mathematical simulations to estimate air pollution levels. These predictions are based on data from emissions sources, chemical transformation models, pollutant dispersion models, topography, roughness, and meteorological data \cite{hoek2017}. Although these methods have been extensively used they come with inherent limitations related to lack of temporal and spatial variation. To overcome the shortcomings, hybrid models (HM) have emerged, blending both spatial and temporal advantages, leading to refined and enhanced estimations \cite{hoek2017, korek2017, tularam2021, oh2021}. \\

While significant advancements have been made in spatiotemporal modelling of air pollution over large-scale domains, such models (multi-country and national domain models) often compromise their accuracy at smaller scales \cite{dehoogh2016, chen2019, chen2020, shen2022}. This gap underscores the need for a comprehensive assessment of outdoor air pollution exposure in smaller domains to understand the health impacts of air pollution exposure, especially during critical stages of development, due to the potential lasting effects of air pollution over the lifespan \cite{selevan2000, wick2010, ghosh2021}. \\

Urban areas add a layer of complexity to air pollution modelling due to complex spatiotemporal patterns that difficult accurate estimation of ground-level air pollutant concentrations \cite{sokhi2022}. In response to these challenges, machine learning (ML) algorithms have been increasingly used, particularly decision trees-based models \cite{liu2022}. This family of models is often used due to its ability to model potential, nonlinear interaction between predictor variables \cite{liu2022treebased}. Among the different ML algorithms, random forest (RF) has emerged as one of the most robust techniques for predicting outdoor air pollution concentrations \cite{chen2019, chen2020, stafoggia2019, stafoggia2020, schneider2020, mila2023}. Furthermore, the COVID-19 pandemic has introduced unprecedented challenges for air pollution modelling and exposure assessment. These challenges encompass changes in air pollution patterns, constraints on data collection due to health and safety restrictions (lockdowns), and the complexity of data integration due to changes in traffic, mobility, and land use dynamics in cities \cite{gonzalez2022, querol2021}. \\

@@

Barcelona (\textbf{\cref{fig1}}), is one of the most populated and dense cities in Spain, with 1,656,725 inhabitants ($\approx16,255.4$ inhabitants/km\textsuperscript{2}). It is located on the Mediterranean coast of the Iberian peninsula, protected by the Collserola mountain range and delimited by two delta rivers (Besòs and LLobregat). The city has a Mediterranean climate, the wind patterns are dominated by a breeze that blows from the sea during the daytime and from the land during night-time. Barcelona is one of the cities in Europe with the highest vehicle traffic density (6,000 vehicles/km\textsuperscript{2}) \cite{casallas2018}. Air pollution measurements were conducted during two separate monitoring campaigns. The first, named the "BiSC-home" campaign, took place at the home addresses of BiSC participants and involved measuring only NO\textsubscript{2} levels for one week during the first, and third trimester of pregnancy between 2018-2021. The second campaign was carried out in accordance with the ESCAPE guidelines, named "BiSCAPE", and included four distinct measurement periods: three in the winter, summer, and autumn of 2021, and a final one in the winter of 2022. During these periods, levels of NO\textsubscript{2}, PM\textsubscript{2.5}, and BC were measured at 34 locations across Barcelona Metropolitan Area (Barcelona, Esplugues de Llobregat, and L'Hospitalet).These monitoring campaigns were used in the development of the LUR and HM.

@@

@@

LUR models for NO\textsubscript{2}  were developed using only data from the locations that had measured home-outdoor (BiSC-home campaign) levels for both sampling campaigns (i.e., we did not use data from participants who had data available for only one campaign, either first or third trimester). Almost half of the sites were measured before the COVID-19 pandemic started and half of the sites were during the pandemic period and its resulting restrictions. For PM\textsubscript{2.5}, BC  and PM\textsubscript{2.5} constituents models were developed using BiSCAPE data from the 34 sites. For the development of the LUR models, spatial variables that better represent the air pollutant concentration distribution across the study area were created. Following the ESCAPE guidelines \cite{eeftens2012, beelen2013},  predictor variables were selected from three broad categories: traffic, land use, and urban configuration. A wide variety of buffer sizes were created, i.e, 25m, 50m, 100m, 300m, 500m, and 1000m to capture the area-specific predictor variable at every single location. Predictor variables were calculated using specific coordinates for each site. We applied a supervised stepwise process in which we sequentially added a predictor variable to a linear regression model using pollutant concentrations as the independent variable to obtain the best possible model \cite{eeftens2012, beelen2013}. This procedure was followed to build all the specific pollutant models. Previously to model development, we specified the direction expected of the effect that each predictor has in the model. For all predictor variables, univariate regression models with annual average air pollution concentration (NO\textsubscript{2}, PM\textsubscript{2.5}, BC, and PM\textsubscript{2.5} constituents (Fe, Cu, Zn)) were conducted, models with the highest adjusted explained variance (adjusted R\textsuperscript{2}) and the right direction of the effect were selected. We then repeated the process using the previous model and all the remaining predictor variables. Details of the LUR modelling and variables used can be found in the \textbf{Table S1}.

@@

\subsubsection{LUR model evaluation}
Performance of the LUR models was assessed using \textit{k-fold cross-validation}. Specifically, we employed a 10-fold cross-validation technique. In this approach, the data was split into ten distinct sub-samples. For each iteration, the model was trained on k-1 folds and then validated using the leftover fold \cite{ziegel2003}. For every fold, we computed metrics such as \(R^2\), the root-mean-square-error (RMSE), mean bias (MB), and the correlation coefficient (\(r\)). After each iteration, both the error and adjusted \(R\textsuperscript{2}\) were recorded. This process was repeated \(k-times\), and the average of these measures was reported as the indicator of the model validity. Details from the metrics used for model validation can be found in the \textbf{Supplementary material X}.\\

@@

\subsection{Dispersion model evaluation}
 Briefly, once input data has been introduced to the ADMS Urban model, the calculation of NO\textsubscript{2}, PM\textsubscript{2.5}, and BC concentration levels has been carried out. Dispersion models were evaluated by comparing the yielded results from the ADM-Urban model with the real values measured by the air quality stations (XVPCA), including Ciutadella, Eixample, Gracia, Palau Reial, Poblenou, Sants, among others. The closer these values are, the better is the air quality modelling.  Details from the metrics used for dispersion model validation and the XVPCA stations locations can be found in the \textbf{supplementary material X}.\vspace{0.5cm}

@@

%%%%%%%%%%%%%%%%%%%%%%%%
%%% --- Results --- %%%
%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

@@

%%% --- Figure. Comparison/agreement between moodel predictions estimates --- %%%
% Put the figure text closer to the Figure 3

% Put the figure text closer to the Figure 3
\captionsetup[figure]{skip=6pt}
% We add the figure with the study domain 
\begin{figure}[!htb]
\includegraphics[width=1.0\textwidth]{figures/HEX_all.png}
\caption{Comparison/agreement between outdoor air pollution model estimates \textit{NO$_2$}, \textit{PM$_{2.5}$}, \textit{BC}.}
\end{figure}
